\bf Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-03-27 paper_original.tex
Added author's suggestion

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{agsm}
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Visualising How Non-linear Dimension Reduction Warps Your Data},
  pdfauthor={Jayani P.G. Lakshika; Dianne Cook; Paul Harrison; Michael Lydeamore; Thiyanga S. Talagala},
  pdfkeywords={high-dimensional data, dimension
reduction, triangulation, hexagonal binning, low-dimensional
manifold, manifold learning, tour, data vizualization},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\section{Introduction}\label{sec-intro}

@@

\centering{

\includegraphics[width=1\textwidth,height=\textheight]{paper_original_files/figure-pdf/fig-nldervisUMAP-1.pdf}

@@

\end{figure}%

\section{Background}\label{sec-background}

\subsection{Dimension Reduction}\label{dimension-reduction}

@@

\subsubsection{Non-linear dimension reduction
techniques}\label{non-linear-dimension-reduction-techniques}

@@

\centering{

\includegraphics[width=1\textwidth,height=\textheight]{paper_original_files/figure-pdf/fig-nldervis-1.pdf}

@@

\end{figure}%

\subsection{Linear overviews using
tours}\label{linear-overviews-using-tours}

@@

\section{Methodology}\label{sec-methods}

@@

\centering{

\includegraphics[width=1\textwidth,height=1\textheight]{figures/workflow.png}

@@

\end{figure}%

\subsection{Preprocessing steps}\label{preprocessing-steps}

@@

\subsection{Constructing the 2D model}\label{sec-construct2d}

@@

\begin{equation}\phantomsection\label{eq-equation1}{
 ar = \frac{r_2}{r_1}
}\end{equation}

\begin{equation}\phantomsection\label{eq-equation2}{
 hr = \frac{hb}{wb}
}\end{equation}

\begin{equation}\phantomsection\label{eq-equation3}{
 y_{max} = \left\lceil\frac{ar}{hr}\right\rceil * hr
}\end{equation}

@@

\begin{equation}\phantomsection\label{eq-equation11}{
 h = \sqrt{3} * s
}\end{equation}

\begin{equation}\phantomsection\label{eq-equation12}{
 b_1 = \frac{r_1 + buffer_{x}}{h}
}\end{equation}

\begin{equation}\phantomsection\label{eq-equation13}{
  v = 1.5 * s
}\end{equation}

\begin{equation}\phantomsection\label{eq-equation14}{
 b_2 = \frac{r_2 + buffer_{y}}{v}
}\end{equation}

@@

\begin{equation}\phantomsection\label{eq-equation15}{
d_x = \frac{h}{2}
}\end{equation}

\begin{equation}\phantomsection\label{eq-equation16}{
d_y = \frac{2* v * 1.15}{\sqrt{3}} 
}\end{equation}

@@

\centering{

\includegraphics[width=1\textwidth,height=1\textheight]{figures/hex_coord.png}

@@

\end{figure}%

@@

\begin{equation}\phantomsection\label{eq-equation4}{
C_k^{(2)} = (C_{ky_1}, C_{ky_2}) = \left(\frac{\sum_{i=1}^{6} h^{k}x_{i}}{6}, \frac{\sum_{i=1}^{6} h^{k}y_{i}}{6}\right).
}\end{equation}

@@

\begin{equation}\phantomsection\label{eq-equation5}{
C_k^{(2)} = (C_{ky_1}, C_{ky_2}) = \left(\frac{1}{n_k} \sum_{i=1}^{n_k} y_{1i}, \frac{1}{n_k} \sum_{i=1}^{n_k} y_{2i}\right),
}\end{equation}

@@

\subsection{Lifting the model into high
dimensions}\label{lifting-the-model-into-high-dimensions}

\subsubsection{Lifting the triangular mesh points into high
dimensions}\label{lifting-the-triangular-mesh-points-into-high-dimensions}

@@

\begin{equation}\phantomsection\label{eq-equation6}{
C_k^{(p)} = (C_{kx_1}, ..., C_{kx_p}) = \left(\frac{1}{n_k} \sum_{i=1}^{n_k} x_{1i}, \frac{1}{n_k} \sum_{i=1}^{n_k} x_{2i}, \dots ,\frac{1}{n_k} \sum_{i=1}^{n_k} x_{pi}\right).
}\end{equation}

@@

\begin{equation}\phantomsection\label{eq-equation7}{
v(C_k^{(2)}) = C_k^{(p)}.
}\end{equation}

@@

\subsubsection{Lifting the 2D triangular mesh into high
dimensions}\label{lifting-the-2d-triangular-mesh-into-high-dimensions}

@@

\centering{

\includegraphics{paper_original_files/figure-pdf/fig-wkhighD-1.pdf}

@@

\end{figure}%

\subsection{Tuning the model}\label{tuning-the-model}

@@

\subsubsection{Total number of bins}\label{total-number-of-bins}

@@

\centering{

\includegraphics{paper_original_files/figure-pdf/fig-binsize-1.pdf}

@@

\end{figure}%

\begin{equation}\phantomsection\label{eq-equation8}{
b = b_1 \times b_2
}\end{equation}

@@

\centering{

\includegraphics{paper_original_files/figure-pdf/fig-diagnosticpltScurve-1.pdf}

@@

\end{figure}%

\subsubsection{Benchmark value to remove low-density
hexagons}\label{benchmark-value-to-remove-low-density-hexagons}

@@

\begin{equation}\phantomsection\label{eq-equationp2}{
\text{standard count} = \frac{\text{count}}{\text{max count}} 
}\end{equation}

@@

\centering{

\includegraphics{paper_original_files/figure-pdf/fig-stdctsScurve-1.pdf}

@@

\centering{

\includegraphics{paper_original_files/figure-pdf/fig-diagnosticpltScurvelwd-1.pdf}

@@

\centering{

\includegraphics{paper_original_files/figure-pdf/fig-rmlowdenshex-1.pdf}

@@

\end{figure}%

\subsubsection{Benchmark value for removing long
edges}\label{benchmark-value-for-removing-long-edges}

@@

\centering{

\includegraphics{paper_original_files/figure-pdf/fig-distScurve-1.pdf}

@@

\centering{

\includegraphics{paper_original_files/figure-pdf/fig-modelScurve-1.pdf}

@@

\end{figure}%

\subsubsection{Starting point of the hexagonal
grid}\label{starting-point-of-the-hexagonal-grid}

@@

\centering{

\includegraphics{paper_original_files/figure-pdf/fig-mseScurvestartpoint-1.pdf}

@@

\centering{

\includegraphics{paper_original_files/figure-pdf/fig-scurveshifthexgridsexp-1.pdf}

@@

\caption{\label{fig-scurveshifthexgridsexp}(a) Visualization of bin
centroids and (b) hexagonal grid, showcasing the configuration before
(colored in gray) and after (colored in light orange) shifting the
starting point by an identical amount applied in both x and y
directions, with a shift value of \(0.537285\).}

\end{figure}%

@@

\centering{

\includegraphics{paper_original_files/figure-pdf/fig-scurveshifthexgrids-1.pdf}

@@

\caption{\label{fig-scurveshifthexgrids}Hexbin plots shows the
distribution of points before and after shifting the starting point of
the hexagonal grid generated for UMAP applied to the training S-curve
dataset. (a) Hexbin plot before shifting, and (b) Hexbin plot after
shifting (identical shift applied in both x and y directions, with a
shift value of \(0.537285\)).}

\end{figure}%

\subsection{Model summaries}\label{sec-summary}

\subsubsection{Predicted values and
residuals}\label{predicted-values-and-residuals}

@@

\begin{equation}\phantomsection\label{eq-equation10}{
\text{Error} = \sum_{j = 1}^{n}\sum_{i = 1}^{p} (x_{ij} - C_{x_ij})^2
}\end{equation}

@@

\subsubsection{Goodness of fit statistics}\label{sec-goodfit}

@@

\begin{equation}\phantomsection\label{eq-equation9}{
\text{MSE} = \sum_{j = 1}^{n} \frac{\sum_{i = 1}^{p} (x_{ij} - C_{x_ij})^2}{n}
}\end{equation}

\subsection{Simulated data example}\label{sec-simpleex}

@@

\centering{

\includegraphics[width=1\textwidth,height=\textheight]{paper_original_files/figure-pdf/fig-nldervis5Gau-1.pdf}

@@

\end{figure}%

To investigate which is the best representation to visualize the
original data or all NLDR methods provide equivalent information, we
visualize the model-in-data space. Models from all NLDR methods show
five well-separated clusters (see Figure~\ref{fig-gau1-sc},
Figure~\ref{fig-gau2-sc}, Figure~\ref{fig-gau3-sc},
Figure~\ref{fig-gau4-sc}, and Figure~\ref{fig-gau5-sc}). This suggests
that for the five Gaussian cluster dataset, all NLDR methods effectively
preserve the global structure. tSNE, UMAP, and PHATE display clusters
with varying densities, indicating their ability to capture
within-cluster variation (see Figure~\ref{fig-gau1-sc},
Figure~\ref{fig-gau2-sc}, and Figure~\ref{fig-gau3-sc}). On the other
hand, both TriMAP and PaCMAP show clusters with flat surfaces,
suggesting a failure to capture within-cluster variation (see
Figure~\ref{fig-gau4-sc} and Figure~\ref{fig-gau5-sc}). Therefore,
TriMAP and PaCMAP do not capture the local structure as effectively as
other methods.

@@

\centering{

\includegraphics[width=1\textwidth,height=\textheight]{paper_original_files/figure-pdf/fig-modelfiveGau-1.pdf}

@@

\begin{minipage}{0.33\linewidth}

\centering{

\captionsetup{labelsep=none}\includegraphics{figures/five_gau_clusters/sc_tsne_1.png}

@@

\subcaption{\label{fig-gau1-sc1}}

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}

\centering{

\captionsetup{labelsep=none}\includegraphics{figures/five_gau_clusters/sc_tsne_2.png}

@@

\subcaption{\label{fig-gau1-sc2}}

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}

\centering{

\captionsetup{labelsep=none}\includegraphics{figures/five_gau_clusters/sc_tsne_3.png}

@@

\subcaption{\label{fig-gau1-sc3}}

\end{minipage}%

\caption{\label{fig-gau1-sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with tSNE in high-D,
a video of the tour animation is available at
(\url{https://youtu.be/oQxEb4wRdHI}).}

\end{figure}%

@@

\begin{minipage}{0.33\linewidth}

\centering{

\captionsetup{labelsep=none}\includegraphics{figures/five_gau_clusters/sc_umap_1.png}

@@

\subcaption{\label{fig-gau2-sc1}}

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}

\centering{

\captionsetup{labelsep=none}\includegraphics{figures/five_gau_clusters/sc_umap_2.png}

@@

\subcaption{\label{fig-gau2-sc2}}

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}

\centering{

\captionsetup{labelsep=none}\includegraphics{figures/five_gau_clusters/sc_umap_3.png}

@@

\subcaption{\label{fig-gau2-sc3}}

\end{minipage}%

\caption{\label{fig-gau2-sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with UMAP in high-D,
a video of the tour animation is available at
(\url{https://youtu.be/JW49csPpDx4}).}

\end{figure}%

@@

\begin{minipage}{0.33\linewidth}

\centering{

\captionsetup{labelsep=none}\includegraphics{figures/five_gau_clusters/sc_phate_1.png}

@@

\subcaption{\label{fig-gau3-sc1}}

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}

\centering{

\captionsetup{labelsep=none}\includegraphics{figures/five_gau_clusters/sc_phate_2.png}

@@

\subcaption{\label{fig-gau3-sc2}}

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}

\centering{

\captionsetup{labelsep=none}\includegraphics{figures/five_gau_clusters/sc_phate_3.png}

@@

\subcaption{\label{fig-gau3-sc3}}

\end{minipage}%

\caption{\label{fig-gau3-sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with PHATE in
high-D, a video of the tour animation is available at
(\url{https://youtu.be/xqhRODt8Z-o}).}

\end{figure}%

@@

\begin{minipage}{0.33\linewidth}

\centering{

\captionsetup{labelsep=none}\includegraphics{figures/five_gau_clusters/sc_trimap_1.png}

@@

\subcaption{\label{fig-gau4-sc1}}

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}

\centering{

\captionsetup{labelsep=none}\includegraphics{figures/five_gau_clusters/sc_trimap_2.png}

@@

\subcaption{\label{fig-gau4-sc2}}

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}

\centering{

\captionsetup{labelsep=none}\includegraphics{figures/five_gau_clusters/sc_trimap_3.png}

@@

\subcaption{\label{fig-gau4-sc3}}

\end{minipage}%

\caption{\label{fig-gau4-sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with TriMAP in
high-D, a video of the tour animation is available at
(\url{https://youtu.be/X-uHSB4VoZ0}).}

\end{figure}%

@@

\begin{minipage}{0.33\linewidth}

\centering{

\captionsetup{labelsep=none}\includegraphics{figures/five_gau_clusters/sc_pacmap_1.png}

@@

\subcaption{\label{fig-gau5-sc1}}

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}

\centering{

\captionsetup{labelsep=none}\includegraphics{figures/five_gau_clusters/sc_pacmap_2.png}

@@

\subcaption{\label{fig-gau5-sc2}}

\end{minipage}%
%
\begin{minipage}{0.33\linewidth}

\centering{

\captionsetup{labelsep=none}\includegraphics{figures/five_gau_clusters/sc_pacmap_3.png}

@@

\subcaption{\label{fig-gau5-sc3}}

\end{minipage}%

\caption{\label{fig-gau5-sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with PaCMAP in
high-D, a video of the tour animation is available at
(\url{https://youtu.be/UxvU8HQd7JM}).}

\end{figure}%

\section{Applications}\label{sec-applications}

\subsection{Single-cell RNA-seq data of
human}\label{single-cell-rna-seq-data-of-human}

@@

\centering{

\includegraphics{paper_original_files/figure-pdf/fig-umapauthor-1.pdf}

>>>>

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{agsm}
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Visualising How Non-linear Dimension Reduction Warps Your Data},
  pdfauthor={Jayani P.G. Lakshika; Dianne Cook; Paul Harrison; Michael Lydeamore; Thiyanga S. Talagala},
  pdfkeywords={high-dimensional data, dimension
reduction, triangulation, hexagonal binning, low-dimensional
manifold, manifold learning, tour, data vizualization},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, sharp corners, enhanced, interior hidden, boxrule=0pt, frame hidden, breakable]}{\end{tcolorbox}}\fi

\hypertarget{sec-intro}{%
\section{Introduction}\label{sec-intro}}

@@

{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_original_files/figure-pdf/fig-nldervisUMAP-1.pdf}

@@

\end{figure}

\hypertarget{sec-background}{%
\section{Background}\label{sec-background}}

\hypertarget{dimension-reduction}{%
\subsection{Dimension Reduction}\label{dimension-reduction}}

@@

\hypertarget{non-linear-dimension-reduction-techniques}{%
\subsubsection{Non-linear dimension reduction
techniques}\label{non-linear-dimension-reduction-techniques}}

@@

{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_original_files/figure-pdf/fig-nldervis-1.pdf}

@@

\end{figure}

\hypertarget{linear-overviews-using-tours}{%
\subsection{Linear overviews using
tours}\label{linear-overviews-using-tours}}

@@

\hypertarget{sec-methods}{%
\section{Methodology}\label{sec-methods}}

@@

{\centering \includegraphics[width=1\textwidth,height=1\textheight]{figures/workflow.png}

@@

\end{figure}

\hypertarget{preprocessing-steps}{%
\subsection{Preprocessing steps}\label{preprocessing-steps}}

@@

\hypertarget{sec-construct2d}{%
\subsection{Constructing the 2D model}\label{sec-construct2d}}

@@

\begin{equation}\protect\hypertarget{eq-equation1}{}{
 ar = \frac{r_2}{r_1}
}\label{eq-equation1}\end{equation}

\begin{equation}\protect\hypertarget{eq-equation2}{}{
 hr = \frac{hb}{wb}
}\label{eq-equation2}\end{equation}

\begin{equation}\protect\hypertarget{eq-equation3}{}{
 y_{max} = \left\lceil\frac{ar}{hr}\right\rceil * hr
}\label{eq-equation3}\end{equation}

@@

\begin{equation}\protect\hypertarget{eq-equation11}{}{
 h = \sqrt{3} * s
}\label{eq-equation11}\end{equation}

\begin{equation}\protect\hypertarget{eq-equation12}{}{
 b_1 = \frac{r_1 + buffer_{x}}{h}
}\label{eq-equation12}\end{equation}

\begin{equation}\protect\hypertarget{eq-equation13}{}{
  v = 1.5 * s
}\label{eq-equation13}\end{equation}

\begin{equation}\protect\hypertarget{eq-equation14}{}{
 b_2 = \frac{r_2 + buffer_{y}}{v}
}\label{eq-equation14}\end{equation}

@@

\begin{equation}\protect\hypertarget{eq-equation15}{}{
d_x = \frac{h}{2}
}\label{eq-equation15}\end{equation}

\begin{equation}\protect\hypertarget{eq-equation16}{}{
d_y = \frac{2* v * 1.15}{\sqrt{3}} 
}\label{eq-equation16}\end{equation}

@@

{\centering \includegraphics[width=1\textwidth,height=1\textheight]{figures/hex_coord.png}

@@

\end{figure}

@@

\begin{equation}\protect\hypertarget{eq-equation4}{}{
C_k^{(2)} = (C_{ky_1}, C_{ky_2}) = \left(\frac{\sum_{i=1}^{6} h^{k}x_{i}}{6}, \frac{\sum_{i=1}^{6} h^{k}y_{i}}{6}\right).
}\label{eq-equation4}\end{equation}

@@

\begin{equation}\protect\hypertarget{eq-equation5}{}{
C_k^{(2)} = (C_{ky_1}, C_{ky_2}) = \left(\frac{1}{n_k} \sum_{i=1}^{n_k} y_{1i}, \frac{1}{n_k} \sum_{i=1}^{n_k} y_{2i}\right),
}\label{eq-equation5}\end{equation}

@@

\hypertarget{lifting-the-model-into-high-dimensions}{%
\subsection{Lifting the model into high
dimensions}\label{lifting-the-model-into-high-dimensions}}

\hypertarget{lifting-the-triangular-mesh-points-into-high-dimensions}{%
\subsubsection{Lifting the triangular mesh points into high
dimensions}\label{lifting-the-triangular-mesh-points-into-high-dimensions}}

@@

\begin{equation}\protect\hypertarget{eq-equation6}{}{
C_k^{(p)} = (C_{kx_1}, ..., C_{kx_p}) = \left(\frac{1}{n_k} \sum_{i=1}^{n_k} x_{1i}, \frac{1}{n_k} \sum_{i=1}^{n_k} x_{2i}, \dots ,\frac{1}{n_k} \sum_{i=1}^{n_k} x_{pi}\right).
}\label{eq-equation6}\end{equation}

@@

\begin{equation}\protect\hypertarget{eq-equation7}{}{
v(C_k^{(2)}) = C_k^{(p)}.
}\label{eq-equation7}\end{equation}

@@

\hypertarget{lifting-the-2d-triangular-mesh-into-high-dimensions}{%
\subsubsection{Lifting the 2D triangular mesh into high
dimensions}\label{lifting-the-2d-triangular-mesh-into-high-dimensions}}

@@

{\centering \includegraphics{paper_original_files/figure-pdf/fig-wkhighD-1.pdf}

@@

\end{figure}

\hypertarget{tuning-the-model}{%
\subsection{Tuning the model}\label{tuning-the-model}}

@@

\hypertarget{total-number-of-bins}{%
\subsubsection{Total number of bins}\label{total-number-of-bins}}

@@

{\centering \includegraphics{paper_original_files/figure-pdf/fig-binsize-1.pdf}

@@

\end{figure}

\begin{equation}\protect\hypertarget{eq-equation8}{}{
b = b_1 \times b_2
}\label{eq-equation8}\end{equation}

@@

{\centering \includegraphics{paper_original_files/figure-pdf/fig-diagnosticpltScurve-1.pdf}

@@

\end{figure}

\hypertarget{benchmark-value-to-remove-low-density-hexagons}{%
\subsubsection{Benchmark value to remove low-density
hexagons}\label{benchmark-value-to-remove-low-density-hexagons}}

@@

\begin{equation}\protect\hypertarget{eq-equationp2}{}{
\text{standard count} = \frac{\text{count}}{\text{max count}} 
}\label{eq-equationp2}\end{equation}

@@

{\centering \includegraphics{paper_original_files/figure-pdf/fig-stdctsScurve-1.pdf}

@@

{\centering \includegraphics{paper_original_files/figure-pdf/fig-diagnosticpltScurvelwd-1.pdf}

@@

{\centering \includegraphics{paper_original_files/figure-pdf/fig-rmlowdenshex-1.pdf}

@@

\end{figure}

\hypertarget{benchmark-value-for-removing-long-edges}{%
\subsubsection{Benchmark value for removing long
edges}\label{benchmark-value-for-removing-long-edges}}

@@

{\centering \includegraphics{paper_original_files/figure-pdf/fig-distScurve-1.pdf}

@@

{\centering \includegraphics{paper_original_files/figure-pdf/fig-modelScurve-1.pdf}

@@

\end{figure}

\hypertarget{starting-point-of-the-hexagonal-grid}{%
\subsubsection{Starting point of the hexagonal
grid}\label{starting-point-of-the-hexagonal-grid}}

@@

{\centering \includegraphics{paper_original_files/figure-pdf/fig-mseScurvestartpoint-1.pdf}

@@

{\centering \includegraphics{paper_original_files/figure-pdf/fig-scurveshifthexgridsexp-1.pdf}

@@

\caption{\label{fig-scurveshifthexgridsexp}(a) Visualization of bin
centroids and (b) hexagonal grid, showcasing the configuration when the
starting point defined at \((-0.113, -0.098)\) (colored in gray) and the
starting point defined at \((-0.025, -0.108)\) (colored in light
orange).}

\end{figure}

@@

{\centering \includegraphics{paper_original_files/figure-pdf/fig-scurveshifthexgrids-1.pdf}

@@

\caption{\label{fig-scurveshifthexgrids}Hexbin plots shows the
distribution of points before and after changing the starting point of
the hexagonal grid generated for UMAP applied to the training S-curve
dataset. (a) Hexbin plot when the starting point defined at
\((-0.113, -0.098)\), and (b) Hexbin plot after changing the starting
point to \((-0.025, -0.108)\).}

\end{figure}

\hypertarget{sec-summary}{%
\subsection{Model summaries}\label{sec-summary}}

\hypertarget{predicted-values-and-residuals}{%
\subsubsection{Predicted values and
residuals}\label{predicted-values-and-residuals}}

@@

\begin{equation}\protect\hypertarget{eq-equation10}{}{
\text{Error} = \sum_{j = 1}^{n}\sum_{i = 1}^{p} (x_{ij} - C_{x_ij})^2
}\label{eq-equation10}\end{equation}

@@

\hypertarget{sec-goodfit}{%
\subsubsection{Goodness of fit statistics}\label{sec-goodfit}}

@@

\begin{equation}\protect\hypertarget{eq-equation9}{}{
\text{MSE} = \sum_{j = 1}^{n} \frac{\sum_{i = 1}^{p} (x_{ij} - C_{x_ij})^2}{n}
}\label{eq-equation9}\end{equation}

\hypertarget{sec-simpleex}{%
\subsection{Simulated data example}\label{sec-simpleex}}

@@

{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_original_files/figure-pdf/fig-nldervis5Gau-1.pdf}

@@

\end{figure}

To investigate which is the best representation to visualize the
original data or all NLDR methods provide equivalent information, we
visualize the model-in-data space. Models from all NLDR methods show
five well-separated clusters (see Figure~\ref{fig-gau1_sc},
Figure~\ref{fig-gau2_sc}, Figure~\ref{fig-gau3_sc},
Figure~\ref{fig-gau4_sc}, and Figure~\ref{fig-gau5_sc}). This suggests
that for the five Gaussian cluster dataset, all NLDR methods effectively
preserve the global structure. tSNE, UMAP, and PHATE display clusters
with varying densities, indicating their ability to capture
within-cluster variation (see Figure~\ref{fig-gau1_sc},
Figure~\ref{fig-gau2_sc}, and Figure~\ref{fig-gau3_sc}). On the other
hand, both TriMAP and PaCMAP show clusters with flat surfaces,
suggesting a failure to capture within-cluster variation (see
Figure~\ref{fig-gau4_sc} and Figure~\ref{fig-gau5_sc}). Therefore,
TriMAP and PaCMAP do not capture the local structure as effectively as
other methods.

@@

{\centering \includegraphics[width=1\textwidth,height=\textheight]{paper_original_files/figure-pdf/fig-modelfiveGau-1.pdf}

@@

\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/five_gau_clusters/sc_tsne_1.png}

@@

@@

\subcaption{\label{fig-gau1_sc1}}
\end{minipage}%
%
\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/five_gau_clusters/sc_tsne_2.png}

@@

\subcaption{\label{fig-gau1_sc2}}
\end{minipage}%
%
\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/five_gau_clusters/sc_tsne_3.png}

\subcaption{\label{fig-gau1_sc3}}
\end{minipage}%

\caption{\label{fig-gau1_sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with tSNE in high-D,
a video of the tour animation is available at
(\url{https://youtu.be/oQxEb4wRdHI}).}

\end{figure}

@@

\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/five_gau_clusters/sc_umap_1.png}

@@

@@

\subcaption{\label{fig-gau2_sc1}}
\end{minipage}%
%
\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/five_gau_clusters/sc_umap_2.png}

@@

\subcaption{\label{fig-gau2_sc2}}
\end{minipage}%
%
\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/five_gau_clusters/sc_umap_3.png}

\subcaption{\label{fig-gau2_sc3}}
\end{minipage}%

\caption{\label{fig-gau2_sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with UMAP in high-D,
a video of the tour animation is available at
(\url{https://youtu.be/JW49csPpDx4}).}

\end{figure}

@@

\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/five_gau_clusters/sc_phate_1.png}

@@

@@

\subcaption{\label{fig-gau3_sc1}}
\end{minipage}%
%
\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/five_gau_clusters/sc_phate_2.png}

@@

\subcaption{\label{fig-gau3_sc2}}
\end{minipage}%
%
\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/five_gau_clusters/sc_phate_3.png}

\subcaption{\label{fig-gau3_sc3}}
\end{minipage}%

\caption{\label{fig-gau3_sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with PHATE in
high-D, a video of the tour animation is available at
(\url{https://youtu.be/xqhRODt8Z-o}).}

\end{figure}

@@

\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/five_gau_clusters/sc_trimap_1.png}

@@

@@

\subcaption{\label{fig-gau4_sc1}}
\end{minipage}%
%
\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/five_gau_clusters/sc_trimap_2.png}

@@

\subcaption{\label{fig-gau4_sc2}}
\end{minipage}%
%
\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/five_gau_clusters/sc_trimap_3.png}

\subcaption{\label{fig-gau4_sc3}}
\end{minipage}%

\caption{\label{fig-gau4_sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with TriMAP in
high-D, a video of the tour animation is available at
(\url{https://youtu.be/X-uHSB4VoZ0}).}

\end{figure}

@@

\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/five_gau_clusters/sc_pacmap_1.png}

@@

@@

\subcaption{\label{fig-gau5_sc1}}
\end{minipage}%
%
\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/five_gau_clusters/sc_pacmap_2.png}

@@

\subcaption{\label{fig-gau5_sc2}}
\end{minipage}%
%
\begin{minipage}[t]{0.33\linewidth}

{\centering

\raisebox{-\height}{

\includegraphics{figures/five_gau_clusters/sc_pacmap_3.png}

\subcaption{\label{fig-gau5_sc3}}
\end{minipage}%

\caption{\label{fig-gau5_sc}Screen shots of the \textbf{langevitour} of
the five Gaussian clusters dataset, shows the model with PaCMAP in
high-D, a video of the tour animation is available at
(\url{https://youtu.be/UxvU8HQd7JM}).}

\end{figure}

\hypertarget{sec-applications}{%
\section{Applications}\label{sec-applications}}

\hypertarget{single-cell-rna-seq-data-of-human}{%
\subsection{Single-cell RNA-seq data of
human}\label{single-cell-rna-seq-data-of-human}}

@@

{\centering \includegraphics{paper_original_files/figure-pdf/fig-umapauthor-1.pdf}
